---
title: "Tutorial on Phonetics and Speech Analysis"
author: 
  - name: <a href='https://www.hugoquene.nl' target='_blank'>Hugo Quené</a>
    affiliation: Utrecht University
date: "Document compiled `r format(Sys.time(), '%d %b %Y %H:%M')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib, pandoc.bib, tphsa.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/TPhSA
description: "Tutorial on Phonetics and Speech Analysis. "
---

# Preface {-}

## Aims {-}

In this tutorial you will learn about **acoustics** (sounds), **phonetics** (speech), and **speech analysis**. 
You will learn the core concepts in these related fields, as well as the necessary practical skills for speech analysis. 
The aim of this tutorial is to provide you with the phonetic insights and skills in speech analysis that you need to succesfully conduct phonetic research in your own project (e.g. paper or thesis). 

### Under construction {-}

This tutorial is a work in progress, resulting from an ongoing revision of an existing tutorial, and meanwhile incorporating other modules and resources. 

The existing (outdated) full tutorial is still available at <https://resources.lab.hum.uu.nl/resources/phonetics/index.html>
.

More details on the origins of this tutorial are provided below. 


## How to use this tutorial {-} 

You will learn the most from this tutorial if you 

(1) read the explanatory texts in this tutorial, 
(2) work through the questions and exercises provided, 
(3) practice in applying your new knowledge hands-on, with the `Praat` computer program (detailed below), and 
(4) re-read the relevant sections from this tutorial and your textbook, with the help of keywords provided per section. 

::: {#question-intro .questionbox}
### Questions {-}

Text blocks such as this one will contain questions or exercises inviting you to engage with the tutorial. You will learn most if you attempt to answer these questions (preferably in writing) *before* you proceed and *before* you take a look at the answer provided. (These questions only work in the HTML version of the tutorial; other versions will just show both the question and answer subsequently.)

#### Question 0.1 {-}

What is sound? 

<details>
  <summary>Answer 0.1</summary>
  <p>Sound is a type of energy that travels through a medium (such as air, water, or solid materials) in the form of waves. These sound waves are created by the vibration of objects, which causes the surrounding particles in the medium to move in a back-and-forth motion. This movement, or vibration, transfers energy through the medium, creating waves of high and low pressure.</p>
</details>

<!-- dummy header ### needed to close tabset -->

:::


## Recommended software {-}

In this tutorial you will work mostly with `Praat` [@Boersma_Weenink_2024]^[The Dutch word _praat_ /ˈpraːt/ means "talk".]. This is a popular open-source program for the analysis of speech, developed by Paul Boersma and David Weenink (both at University of Amsterdam). It can be found on its own website (<https://www.praat.org>), where you will find a wealth of helpful documentation. `Praat` also has extensive `Help` built in, including a full tutorial.  
There is an online forum (<https://groups.io/g/Praat-Users-List>), where users share their knowledge by posting questions and providing answers.

In order to install `Praat` on your computer, go to its webpage at <https://www.praat.org/>, and then proceed to the download page for the operating system of your computer. Follow the installation instructions on the download page for your operating system.  

::: {#tech-layout .praatbox}
### Instructions for using `Praat` {-}

Text blocks such as this one will contain instructions about how to "do" things in `Praat`.

Options in software menus, as well as texts in on-screen buttons, will be shown `in this way`. 
The notation `Main > Sub` means: first choose option `Main` from the main menu, after which a submenu will appear, then choose option `Sub` from the submenu. 
Commands or formulas that you have to type will be shown `in this way` too. (Commands typically need to be terminated with typing `Enter` or `Return` or `␍` or `⏎` -- which however will not be specified in the instructions.)
:::

## Structure of this tutorial {-}

TBA

## Recommended textbooks {#sec:textbooks}

This tutorial is intended to be used in addition to one or more textbook(s) in Phonetics, to which this tutorial will provide additional background knowledge. Some excellent textbooks in Phonetics are those by 
@Rietveld_VanHeuven_2009 (in Dutch), 
@Johnson_2012, 
@Ladefoged_Johnson_2015,  
@Reetz_Jongman_2020, and 
@Zsiga_2024. 


## Details {-}

### License {-}

This work is licensed under the _GNU GPL 3_ license (for details see 
https://www.gnu.org/licenses/gpl-3.0.en.html). 

### Citation {-}

TBA 

### Technical details {-}

TBA

### History {-}

This work is based on an earlier tutorial (2006-2007) titled <cite>Tutorial for self study: basics of phonetics and how to use Praat</cite> by Clizia Welker and Hugo Quené. In turn, that 2007 tutorial was partly based on older texts by Hugo Quené, Denise Bruin and Mirjam Wester (1996-2000); these older texts acknowledged valuable comments and suggestions by Paul Boersma, Olga van Herwijnen, Kim Koppen, Eva Sittig, Joyce Vliegen and Mieke van Wijck. 

The 2007 version of the tutorial was subsequently revised and adapted to the current version using `R Markdown` [@rmarkdown2018] and `bookdown` [@R-bookdown] in [Rstudio](https://www.rstudio.com) by Hugo Quené in 2024.  


```{r bibpackages, class.source='fold-hide', include=FALSE}
# automatically create a bib database for R packages that are used 
# require(lattice)
# require(nnet)
# added 20220203
require(hqmisc)
# require(MASS)
knitr::write_bib( c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

